MLH1 Results Outline

### 2020-03-02 doc/Results.rmd
cleaning up the final draft

In an attempt to localize  the above male specific crossover number evolution within the meiotic pathway we quantified a marker for double strand breaks (DSBs), DMC1, in prophase spermtaocytes. DMC1 foci were scored from a total of `r sum( (DMC1.mean.table %>%  filter(stage == "L") )$ncells)` leptotene and `r sum( (DMC1.mean.table %>%  filter(stage == "Z") )$ncells)` zygotene staged spermatocytes from juvenile mice (12 to 18 days) of three low rec, ( _musculus^KAZ^_ ,  _domesticus^WSB^_ , and _domesticus^G^_) and two high recombining strains ( _musculus^PWD^_ and _molossinus^MSM^_).

@@

The mean number of DMC1 foci per cell is associated with evolution of the number of precursors (DSBs). The two high recombining strains tested have significantly more DMC1 foci compared to low recombining strains in leptoene cells (t.test; p=`r ttest.HighLow.L.pval`, one-way-anova; p=`r format(l.anova.pval, scientific=F)`). However the differences in DMC1 foci were not significant for zygotene cells (later prophase) between the two groups (t.test; p=`r ttest.HighLow.Z.pval`, one-way-anova; p=`r z.anova.pval`). Indicating that DMC1 counts at this stage is more predictive of the downstream crossover number differences.

@@

A limitation of our image analysis algorithm is that not all bivalents per cell are isolated, due to overlaping chromosomts. The range isolation rates per cell in this data set is `r min(algorithm.stat.category$mean_pass.rate)` in _molossinus^MSM^_ male to `r max(algorithm.stat.category$mean_pass.rate)`  _musculus^KAZ^_ female. However, we assume that the isolation of chromsomes within cell images is not biased. From our total set of cell images `r length(unique(new.Curate.FULL.org %>% filter(SC.pass == "1"))$fileName )` chromosome objects were isolated by the image analysis software. After the human curation step, `r length(unique(new.Curate.FULL.org %>% filter(hand.foci.count != ""))$Obj.ID )` single bivalent observations remained. Given the large number of single chromosome observations, we assume that each of the data sets are equally representative of general patterns.

A limitation of these data from cytological stains are the only chromosome which can be accruately identified is the XY in males. The unpaired XY chromosomes have SYCP3 signal along the chromsome axis, but it forms a structure distinct from the synapsed autosomes, which can be excluded in the curationstep. Since it has homologous pair in oocytes, the XX bivalent is indistinguishable from the rest of the autosomes. By phycisal length in Mb, the X is predictied to be the 3rd longest bivalent.

@@

We distill the variance observed in the mean MLH1 foci counts per cell to number of MLH1 foci per bivalent. Because our observations of MLH1 foci are made of pachtyene cells, our chromosome observations are fully synapsed paired homologous chromosomes (4n copies of each chromosomes). Additionally we can not fully distinguish the identity of chromsomes (e.g. Chromosome 1 or Chromosome 2), for these reasons we use the term 'bivalent' instead of chromosome.

Ninety six percent of the pooled dataset of single bivalents (n=`r total.biv.all`) have either one or two crossovers (Figure X). The proportion of 1CO:2CO distinguishes the high and low recombining strains (Figure X). This confirms the intuitive interpretation that high recombination strains are enriched for 2CO bivalents at the expense of 1CO bivalents. In the high recombiningstrains the 2CO proportions are `r chrm.prop.male.high1$n2CO[chrm.prop.male.high1$strain == "SKIVE"] / chrm.prop.male.high1$nbiv[chrm.prop.male.high1$strain == "SKIVE"]` ( _musculus^SKIVE^_ ), `r chrm.prop.male.high1$n2CO[chrm.prop.male.high1$strain == "PWD"] / chrm.prop.male.high1$nbiv[chrm.prop.male.high1$strain == "PWD"]` ( _musculus^PWD^_ ), and `r chrm.prop.male.high1$n2CO[chrm.prop.male.high1$strain == "MSM"] / chrm.prop.male.high1$nbiv[chrm.prop.male.high1$strain == "MSM"]` (_molossinus^MSM^_ ).

While the proportions are significantly different (chi-square-test; p = `r PWD.MSM.2CO.prop.test$p.value`), males _musculus^PWD^_ and _molossinus^MSM^_ have 2CO proportions more similar to each other than strains from the same subspecies ( _musculus^PWD^_ and _musculus^KAZ^_ chi-square test; 2CO p=`r PWD.KAZ.2CO.prop.test$p.value`) and _molossinus^MSM^_ and _molossinus^MOLF^_ chi-square test; 2CO p=`r MSM.MOLF.2CO.prop.test$p.value`).

Using this data set of single bivalents, we focus on aspects of the recombiation landscape to address two main questions 1) Which traits are sexually dimorphic? and 2) Which traits fit distinguish the high and low recombining strains for males?

In our analysis our first step is to test for measurable differences, between sexes or recobination groups in males and if the differences are significant, our second step is to apply the same sets of models applied to the mean MLH1 foci counts to test for effects due to subspecies or strains.

@@

In many mammalian species the SC lengths are longer in females, yet most all of these observations come from species with female biased heterochiasmy. Our sample of house mice provides an opportunity to test if the sex differences in chromatin compaction (SC length) are reversed in the high recombining strains (espicially the high recombining strains with male-biased heterochiasmy; _musculus^PWD^_  and _molossinus^MSM^_). A simple model might predict if SC length is the strongest predictor for number of MLH1 foci per cell males of the high recombining strains would have longer SC length leading to the difference in total MLH1 foci per cell.

@@

For an addtional metric of chromatin compaction, we tested a whole cell measure of SC length using an second image analysis algorithm which converts the SC signal (into single pixel wide skeleton) and provides accurate estimates of the summed SC length of bivalents for single cells (ref Wang). We apply a skeletonizing image analysis algorithm to all MLH1 meiocyte spreads images. Mouse means were calculated from `r length(  (MLH1.merge.Skel %>% filter(species == "M.musculus") %>% filter(subsp != "Cast")  )$Random.Name)` cell wide total SC measures (figure X). (where as `r length(  (MLH1_data %>% filter(species == "M.musculus") %>% filter(subsp != "Cast")  )$Random.Name)` number of cells had MLH1 measures ).

@@

**!! SUMARIZE and remove most of the models pvalues!!**

From the linear models a general pattern that emerges is that sex is a major effect but there are also some subspeices and strain effects, indicating that the SC lengths (degree of chromatin compaction) can evolve in both males and females.  (weaker support in terms of the p values -- but they hint at the musculus and molossinus subspecies having longer sc and WSB SC being espicailly low)

@@

These combined results of chromatoin compaction and SC length, indicate that females have longer SC lengths (chromosome axis), even in strains where males have more MLH1 foci per cell. I the means MLH1 foci per cell and SC area per cell, are considered seperatly, it is as if the males in these strains have more crossovers (more 2CO), but generally less 'space' to place them. <(this suggests that the landcapes between spermatocytes and oocytes -- are different)>

@@

focus on the rec landscape of a major class of chromosomes 1CO and how these landscapes can vary between males and females.

@@

For the Mixed model, sex is the most significant effect (LTR; p =`r lmer.Q1.nrm1CO_M1_results$sex.results`).The random effect of strain is also significant (LRT; p =`r lmer.Q1.nrm1CO_M1_results$rand.pval`). The linear models were used to detect if specific strains differed in the sex differences in the 1CO landscape. While sex was the most significant effect for both M2 and M3 (there were some strain specific effects...)

The male effects were the most significant (glm M2; p =`r summry.Q1.glm_F1pos_M2$coefficients[,4][8]`, glm M3; p = `r summry.Q1.glm_1CO.pos_M3$coefficients[,4][8]`). While the musc subspecies, LEW and KAZ strains had slightly signifcant effects (glm M2; Musc subsp p = `r summry.Q1.glm_F1pos_M2$coefficients[,4][2]`, LEW strain p = `r summry.Q1.glm_F1pos_M2$coefficients[,4][5]`) and (glm M3; LEW strain p = `r summry.Q1.glm_1CO.pos_M3$coefficients[,4][3]`, KAZ strain p = `r summry.Q1.glm_1CO.pos_M3$coefficients[,4][7]`).

<!--
**M1. Mixed Model**
$$mouse \ average \ F1 position ~=~ subsp * sex + rand(strain) + \varepsilon $$

@@

#real.long.bivData

Q1.short.biv_scatr_box   <- ggplot( (real.short.bivData %>% filter(hand.foci.count == 1) ), aes(x=sex, y=PER_Foci_1, color=strain) )+geom_jitter( )+ geom_boxplot(aes(y= PER_Foci_1 ),alpha=.3 ) + scale_color_manual(values=colors_of_strains)+facet_wrap(~subsp, scales = "free")+ylim(c(0,1.1))+ggtitle("Mouse av. \nNormalized 1CO Foci pos \nShort Biv")

@@

Crossover interference is one of the major determinates of the recombination landscape. It generates a distribution of evenly spaced crossover along chromosomes instead of a random and more uniform distribution.

 We isolated `r length(  (Curated_BivData.22 %>% filter(sex == "female") )$Obj.ID )` and 
`r length(  (Curated_BivData.22 %>% filter(sex == "male") )$Obj.ID )` 2CO bivalents in females and males to test describe sex differences in crossover interference. Mean interfocal distances (IFD) were calculated from `r length( unique( (Curated_BivData.22 %>% filter(sex == "female") )$mouse) )`  female and `r length( unique( (Curated_BivData.22 %>% filter(sex == "male") )$mouse) )` male mice. We examined both raw IFD^raw^ and normalized by SC length (IFD^PER^). The mean IFD^raw^ measures of crossover interference as a mechanical force mediated through the SC. We compared the IFD^raw^ to test the hypothesis of interference acting as a mechanical force transmitted through the SC. Previous measures found no sex difference in measures of IFD^raw^ in mice (de Boer et al ).

@@

While the comparisons of the IFD^PER^ metrics can reveal more general recombination landscape patterns while controlling for the underlying differences in chromatin compaction and SC length. Also this metric is closer to measures of interference from linkage maps which are also removed from physical scales in that they measure frequencies of crossovers. Crossover interference is stronger in male specific linkage maps compared to female maps in a variety of species (ref).

The difference in mean IFD^PER^ are highly significant between sexes. The mean IFD^PER^ are significantly long in six of seven strains (MSM t.test; p-value = `r Q1.IFD.PER_t.test_MSM$p.value`), _musculus^SKIVE^_ p-value = `r Q1.IFD.PER_t.test_SKIVE$p.value`), _musculus^PWD^_ p-value = `r Q1.IFD.PER_t.test_PWD$p.value`), _molossinus^MSM^_ p-value =`r Q1.IFD.PER_t.test_MSM$p.value`), _domesticus^LEW^_ p-value = `r Q1.IFD.PER_t.test_LEW$p.value`), _domesticus^G^_ p-value = `r Q1.IFD.PER_t.test_G$p.value`), and _domesticus^WSB^_ (t.test; p-value =`r Q1.IFD.PER_t.test_WSB$p.value`) ). Only _musculus^KAZ^_ did not have a difference in mean IFD^PER^ (t.test; p-value = `r Q1.IFD.PER_t.test_KAZ$p.value`).

<Full dataset (t.test; p = `r Q1.IFD.PER_FULL_t.test$p.value`).>

<!--  finer distribution patterns -->

@@

GENERAL RESTULTS DIFFERENCES between sexes

In examining the IFD^PER^ distributions, we note that 
the female norm IFD are centered at ~50% – but they also seem to have slight an enrichment of short (<25%) normalized IFDs

While the general pattern in the male distributions are enrichment longer IFD^PER^. (however we note that there are slight differences in the IFD^PER^ distributions between the high and low recombining males to be followed up in later section)

<!--
The differences between the sexes is slightly different depending on if the males are in the high rec group. In the low rec strains (Dom + KAZ)...
(both males and females IFD distributions have lower and upper thresholds around 25% and 75%) respectively. however the weight of observations are slightly shifted across the sexes. There's an enrichment of lower IFD^PER^ in females relative to males, while the opposite is true for the higher threshold -- where males have an enrichment of higher/long IFD^PER^ compared to females.

In the three high rec strains -- the distributions have a more striking difference of the lower threshold being much higher in males, aproximately 30% to 40% compared to 25% in females. Also while the female upper threshold is capped at 75%, in males (some strains have higher upper thresholds -- and it's more likely to observe IFD^PER^ above the threshold.

@@

Since the most significant t.test were significantly different across sexes for IFD^PER^ and we follow up with the model framework. For the Mixed models of IFDs, sex is a significant effect for both raw and nrmIFD. (IFD.PER), sex is most significant effect (LRT: p = `r Q1.IFD.PER_M1_lmer.results$sex.results`) subspecies, interaction are slightly significant. <ONLY REPORT FOR PER MEASURES>.  For the normalized values in both M2 and M3, sex is a significant effect, increasing nrm.IFD in males. SKIVE*male is the only other consistently significant effect, which also increases the nrm.IFD measure (p = male-SKIVE p =  ).  When interference is measured in physical SC units, the differences between sexes is low / only slightly significant -leading support that the phyisical measures of interference (in SC units) is conserved within species. While when the variance in chromatin compaction is controlled males have stronger interference (like genetic maps) the impact on the rec landscapes being.

@@

@@

# Q1 Summary, consequenced of sex differences in the recombination landscape

Our results confirm the typical landscape (sex differences). The terminal position of single foci and the greater distance between two foci on the same chromsome will have a consequence of ... The genetic consequence of such a pattern is that larger 'chuncks' a greater proportion of linked sites will segregate together in males compared to females (Veller). (having the effect of fewer novel combinations of site pairs along the total chromosome).

@@

The results from the previous section -- confirm (known sex differences for the 'typical' 
recombination landscape (Sardell Kirkpatrick). The next section is meant to focus on the greater aspect of variation in mean MLH1 counts per cell the high recombining strains ( _musculus^PWD^_ , _musculus^SKIVE^_, and _molossinus^MSM^_) and the rest of the males with the main objective of this section is to test for significant correlations with the evolution of mean MLH1 foci per cell.

In comparing the male specific single bivalent based metrics the first task/level is to tests for significant differences between the high and low recombining groups. The second being to follow up with linear models that test for effects of subspecies and strains.

The literature provides basic predictions for the direction of the relationship the genome wide recombination rate and two aspects of the recombiantion landscape; SC length (positve) and crossover interference (negative). Following this logic we predict i) no strain effects for SC length or crossover interference within _domesticus_, ii) _musculus^PWD^_ will have greater SC length and weaker interference than _musculus^SKIVE^_, which will have longer SC and weaker interference compared to the other _musculus_ strains and iii) _molossinus^MSM^_ will have longer SC and weaker interference compared to _molossinus^MOLF^_.

@@

Our prediction for this section is that the high recombiningstrains will have longer SC lengths proportional to the increase in the mean genome wide recombination rate.

The confounding effects of sex chromosomes are removed for this section that compares across males from different inbred strains. We compared mouse means of three SC based metrics; total SC, mean short bivalent length and mean long bivalent length. The long bivalent data set was isolated in the same manner as the short bivalent data set.

<!-- basic stats
`r length(OG.MLH1.merge.Skel$fileName)` total skelentonize measures were made from all of the MLH1 pachtyene cell images. X observations removed for having extra signal from other cells. `r length(unique( all.remove ) )` measures removed as potential outliers (by examining the category specific plots)-->

<pattern for the single bivalent, pos correlation, The predicted pattern for SC length becomes more nuanced when the data are split up by Chromosome class. Mainly that 1CO are shorter in the high recombiningstrains then low recombining strains. This is likely due to the fact that more physically longer chromosomes have 1CO in the low recombining strains, which pushes the mean SC length up. Where as in the high recombiningstrains, physically longer chromosomes are more likely to be in the 2CO group.  This supports a general pattern of tighter clustering of SC lengths across chromosome classes in the high recombininggroup. They have a lower probability of 2COs below a certain SC length threshold.>

@@

There is a positive correlation between mean MLH1 foci per cell and total SC, however the variation in mean total SC does not completely predict the high and low recombining strains (Figure X). While for the total data set of mean total SC, the high recombiningstrains have significantly  (ttest; p = `r Q2tot.sc_t.test_High.Low$p.value`) when the high and low recombining strains are compared within subspecies only the _molossinus_ strains have significant difference  (ttest mol p = `r Q2tot.sc_t.test_High.Low_Mol$p.value`), while the _musculus_ strains do not have significant differences (ttest; musc p= `r Q2tot.sc_t.test_High.Low_Musc$p.value`).

Additionally, the means for the reduced bivalent datasets, short and long bivalents, are not significantly different between the high and low recombining strains (ttest; short p = `r Q2short.biv_t.test_High.Low$p.value` and long p = `r Q2long.biv_t.test_High.Low$p.value`).

@@

<!-- models 
**Linear model M1 ** $$mouse \ average \ SC metric ~=~ subsp * strain + \varepsilon $$
**Linear model M2 ** $$mouse \ average \ SC metric ~=~ strain + \varepsilon $$-->

@@

We follow up the general pattern of a weak signal of high recombiningstrains having longer SC lengths by running the model framework. For total SC in M1 (glm including subsp), the two subspecies effects are significant (glm; p= _Musculus_ `r sum.Q2.glm_mus.av_totSC_M1[,4][2]` and _Molossinus_ p= `r sum.Q2.glm_mus.av_totSC_M1[,4][3]`). (in the models using the short and long reduced datasets, (significant effects are /range from musc and strain MSM, (glm; p < 0.05). In the short bivalent data set, analyzed by M1, both subspecies are sig (glm; p= _Musculus_ `r sum.glmQ2_short.biv_M1$coefficients[,4][2]` and _Molossinus_ p= `r sum.glmQ2_short.biv_M1$coefficients[,4][3]`). While the models for the reduced dta set suggest that the subspecies with the high recombining strains have evolved longer SC lengths -- But the differences in // but the longer SC lengths do ot cleanily predict the higher recombining strains (there are longer SC in ). To some extent the SC lenth evolution is decoupled from the mean MLH1 foci per cell (since some strains have longer sc - but 'lower' recombination rates)

<!--
long bivalent set in M1 -- 
Musc `r sum.glmQ2_long.biv_M1$coefficients[,4][2]` 
MSM strain `r sum.glmQ2_long.biv_M1$coefficients[,4][7]`
<And the strain effect for G is significant (glm; G p =`r sum.glmQ2_short.biv_M1$coefficients[,4][4]`).>

@@

Unlike total SC length and single foci landscapes, our metrics of interference from 2COs are significant predictors of high recombining strains (gwRR evolution).

>>>>

In an attempt to localize  the above male specific crossover number evolution within the meiotic pathway we quantified a marker for double strand breaks (DSBs), DMC1, in prophase spermatocytes. DMC1 foci were scored from a total of `r sum( (DMC1.mean.table %>%  filter(stage == "L") )$ncells)` leptotene and `r sum( (DMC1.mean.table %>%  filter(stage == "Z") )$ncells)` zygotene staged spermatocytes from juvenile mice (12 to 18 days) of three low rec, ( _musculus^KAZ^_ ,  _domesticus^WSB^_ , and _domesticus^G^_) and two high recombining strains ( _musculus^PWD^_ and _molossinus^MSM^_).

@@

The mean number of DMC1 foci per cell is associated with evolution of the number of precursors (DSBs). The two high recombining strains tested have significantly more DMC1 foci compared to low recombining strains in leptotene cells (t.test; p=`r ttest.HighLow.L.pval`, one-way-anova; p=`r format(l.anova.pval, scientific=F)`). However the differences in DMC1 foci were not significant for zygotene cells (later prophase) between the two groups (t.test; p=`r ttest.HighLow.Z.pval`, one-way-anova; p=`r z.anova.pval`). Indicating that DMC1 counts at this stage is more predictive of the downstream crossover number differences.

@@

A limitation of our image analysis algorithm is that not all bivalents per cell are isolated, due to overlapping  chromosomes. The range isolation rates per cell in this data set is `r min(algorithm.stat.category$mean_pass.rate)` in _molossinus^MSM^_ male to `r max(algorithm.stat.category$mean_pass.rate)`  _musculus^KAZ^_ female. However, we assume that the isolation of chromsomes within cell images is not biased. From our total set of cell images `r length(unique(new.Curate.FULL.org %>% filter(SC.pass == "1"))$fileName )` chromosome objects were isolated by the image analysis software. After the human curation step, `r length(unique(new.Curate.FULL.org %>% filter(hand.foci.count != ""))$Obj.ID )` single bivalent observations remained. Given the large number of single chromosome observations, we assume that each of the data sets are equally representative of general patterns.

A limitation of these data from cytological stains are the only chromosome which can be accurately identified is the XY in males. The unpaired XY chromosomes have SYCP3 signal along the chromosome axis, but it forms a structure distinct from the synapsed autosomes, which can be excluded in the curation step. Since it has homologous pair in oocytes, the XX bivalent is indistinguishable from the rest of the autosomes. By physical length in Mb, the X is predicted to be the 3rd longest bivalent.

@@

We distill the variance observed in the mean MLH1 foci counts per cell to number of MLH1 foci per bivalent. Because our observations of MLH1 foci are made of pachytene cells, our chromosome observations are fully synapsed paired homologous chromosomes (4n copies of each chromosomes). Additionally we can not fully distinguish the identity of chromosomes (e.g. Chromosome 1 or Chromosome 2), for these reasons we use the term 'bivalent' instead of chromosome.

Ninety six percent of the pooled dataset of single bivalents (n=`r total.biv.all`) have either one or two crossovers (Figure X). The proportion of 1CO:2CO distinguishes the high and low recombining strains (Figure X). This confirms the intuitive interpretation that high recombination strains are enriched for 2CO bivalents at the expense of 1CO bivalents. In the high recombining strains the 2CO proportions are `r chrm.prop.male.high1$n2CO[chrm.prop.male.high1$strain == "SKIVE"] / chrm.prop.male.high1$nbiv[chrm.prop.male.high1$strain == "SKIVE"]` ( _musculus^SKIVE^_ ), `r chrm.prop.male.high1$n2CO[chrm.prop.male.high1$strain == "PWD"] / chrm.prop.male.high1$nbiv[chrm.prop.male.high1$strain == "PWD"]` ( _musculus^PWD^_ ), and `r chrm.prop.male.high1$n2CO[chrm.prop.male.high1$strain == "MSM"] / chrm.prop.male.high1$nbiv[chrm.prop.male.high1$strain == "MSM"]` (_molossinus^MSM^_ ).

While the proportions are significantly different (chi-square-test; p = `r PWD.MSM.2CO.prop.test$p.value`), males _musculus^PWD^_ and _molossinus^MSM^_ have 2CO proportions more similar to each other than strains from the same subspecies ( _musculus^PWD^_ and _musculus^KAZ^_ chi-square test; 2CO p=`r PWD.KAZ.2CO.prop.test$p.value`) and _molossinus^MSM^_ and _molossinus^MOLF^_ chi-square test; 2CO p=`r MSM.MOLF.2CO.prop.test$p.value`). Using this data set of single bivalents, we focus on aspects of the recombination landscape to address two main questions 1) Which traits are sexually dimorphic? and 2) Which traits fit distinguish the high and low recombining strains for males?

In our analysis our first step is to test for measurable differences, between sexes or recombination groups in males and if the differences are significant, our second step is to apply the same sets of models applied to the mean MLH1 foci counts to test for effects due to subspecies or strains.

@@

In many mammalian species the SC lengths are longer in females, yet the majority of these observations come from species with female biased heterochiasmy. Our sample of house mice provides an opportunity to test if the sex differences in chromatin compaction (SC length) are reversed in the high recombining strains with male-biased heterochiasmy; _musculus^PWD^_  and _molossinus^MSM^_). If SC length is the strongest predictor for number of MLH1 foci per cell, males of the high recombining strains would have longer SC length leading to the difference in total MLH1 foci per cell.

@@

For an additional  metric of chromatin compaction, we tested a whole cell measure of SC length using a second image analysis algorithm which converts the SC signal into single pixel wide trace of the SC shapes and provides accurate estimates of the summed SC length of bivalents for single cells (ref Wang). We apply the skeletonizing algorithm to all MLH1 meiocyte images. Mouse means were calculated from `r length(  (MLH1.merge.Skel %>% filter(species == "M.musculus") %>% filter(subsp != "Cast")  )$Random.Name)` cell wide total SC measures (figure X). (where as `r length(  (MLH1_data %>% filter(species == "M.musculus") %>% filter(subsp != "Cast")  )$Random.Name)` number of cells had MLH1 measures ).

@@

@@

Longer Sc compaction in females is supporte by the model framework, with sex being the most consistant significant effect (LTR and glm; p < 0.05). Additionally there are some significant subspecies and strain effects indicating that the SC lengths can evolve within strains and subspecies (LTR and glm; p < 0.05).

These combined results of total SC per cell and reduced single bivalent SC metrics, indicate that females have longer SC lengths (chromosome axis), even in strains where males have more MLH1 foci per cell. I the means MLH1 foci per cell and SC area per cell, are considered seperatly, it is as if the males in these strains have more crossovers (more 2CO), but generally less 'space' to place them.

@@

For most of the house mouse strains, 1COs are the major class of chromosomes observed across cells. There are known sex differences in the relative placement along chromsomes this section tests if there has been evolution in this aspect of the recombination landscape.

@@

The differences in 1CO landscapes due to sexes is confirmed by the model framework, with sex is the most significant effect in the mixed model (M1) (LTR; p =`r lmer.Q1.nrm1CO_M1_results$sex.results`) and the two linear models (glm M2; p =`r summry.Q1.glm_F1pos_M2$coefficients[,4][8]`, glm M3; p = `r summry.Q1.glm_1CO.pos_M3$coefficients[,4][8]`).

<!--
not addressing the strain and subspecies effects
The random effect of strain is also significant (LRT; p =`r lmer.Q1.nrm1CO_M1_results$rand.pval`).

While the musc subspecies, LEW and KAZ strains had slightly significant effects (glm M2; Musc subsp p = `r summry.Q1.glm_F1pos_M2$coefficients[,4][2]`, LEW strain p = `r summry.Q1.glm_F1pos_M2$coefficients[,4][5]`) and (glm M3; LEW strain p = `r summry.Q1.glm_1CO.pos_M3$coefficients[,4][3]`, KAZ strain p = `r summry.Q1.glm_1CO.pos_M3$coefficients[,4][7]`).

**M1. Mixed Model**
$$mouse \ average \ F1 position ~=~ subsp * sex + rand(strain) + \varepsilon $$

@@

#real.long.bivData
Q1.short.biv_scatr_box   <- ggplot( (real.short.bivData %>% filter(hand.foci.count == 1) ), aes(x=sex, y=PER_Foci_1, color=strain) )+geom_jitter( )+ geom_boxplot(aes(y= PER_Foci_1 ),alpha=.3 ) + scale_color_manual(values=colors_of_strains)+facet_wrap(~subsp, scales = "free")+ylim(c(0,1.1))+ggtitle("Mouse av. \nNormalized 1CO Foci pos \nShort Biv")

@@

Crossover interference is one of the major determinates of the recombination landscape. It generates a distribution of evenly spaced crossover along chromosomes instead of a random and more uniform distribution. To quantifiy how crossover interference differs between the typical recombination landscapes of males and females, we isolated `r length(  (Curated_BivData.22 %>% filter(sex == "female") )$Obj.ID )` and 
`r length(  (Curated_BivData.22 %>% filter(sex == "male") )$Obj.ID )` 2CO bivalents in females and males respectively. Mean interfocal distances (IFD) were calculated from `r length( unique( (Curated_BivData.22 %>% filter(sex == "female") )$mouse) )`  female and `r length( unique( (Curated_BivData.22 %>% filter(sex == "male") )$mouse) )` male mice. We examined both raw IFD^raw^ and normalized by SC length (IFD^norm^). The mean IFD^raw^ measures of crossover interference as a mechanical force mediated through the SC. We compared the IFD^raw^ to test the hypothesis of interference acting as a mechanical force transmitted through the SC. Previous measures found no sex difference in measures of IFD^raw^ in mice (de Boer et al 2004).

@@

While the comparisons of the IFD^norm^ metrics can reveal more general recombination landscape patterns while controlling for the underlying differences in chromatin compaction and SC length. This metric is closer to measures of interference from linkage maps which are also removed from physical scales in that they measure frequencies of crossovers. Crossover interference is stronger in male specific linkage maps compared to female maps in a variety of species (ref).

The difference in mean IFD^norm^ are highly significant between sexes. The mean IFD^norm^ are significantly long in six of seven strains (MSM t.test; p-value = `r Q1.IFD.PER_t.test_MSM$p.value`), _musculus^SKIVE^_ p-value = `r Q1.IFD.PER_t.test_SKIVE$p.value`), _musculus^PWD^_ p-value = `r Q1.IFD.PER_t.test_PWD$p.value`), _molossinus^MSM^_ p-value =`r Q1.IFD.PER_t.test_MSM$p.value`), _domesticus^LEW^_ p-value = `r Q1.IFD.PER_t.test_LEW$p.value`), _domesticus^G^_ p-value = `r Q1.IFD.PER_t.test_G$p.value`), and _domesticus^WSB^_ (t.test; p-value =`r Q1.IFD.PER_t.test_WSB$p.value`) ). Only _musculus^KAZ^_ did not have a difference in mean IFD^norm^ (t.test; p-value = `r Q1.IFD.PER_t.test_KAZ$p.value`).

<!--  finer distribution patterns <Full dataset (t.test; p = `r Q1.IFD.PER_FULL_t.test$p.value`).> -->

@@

In examining the IFD^norm^ distributions, we note that the female IFD^norm^ observations are centered at approximately 50% and have slight an enrichment of short (<25%) IFD^norm^ observations. Comparitively, the IDF^norm^ distributions in males are enriched for longer IFD^norm^.

<!--
The differences between the sexes is slightly different depending on if the males are in the high rec group. In the low rec strains (Dom + KAZ)...
(both males and females IFD distributions have lower and upper thresholds around 25% and 75%) respectively. however the weight of observations are slightly shifted across the sexes. There's an enrichment of lower IFD^norm^ in females relative to males, while the opposite is true for the higher threshold -- where males have an enrichment of higher/long IFD^norm^ compared to females.

In the three high rec strains -- the distributions have a more striking difference of the lower threshold being much higher in males, aproximately 30% to 40% compared to 25% in females. Also while the female upper threshold is capped at 75%, in males (some strains have higher upper thresholds -- and it's more likely to observe IFD^norm^ above the threshold.

@@

The three sets of models support the result of stronger interference in males, sex was the most significant significant effect (LRT: p = `r Q1.IFD.PER_M1_lmer.results$sex.results`). When interference is measured in physical SC units (IFD^raw^), the differences between sexes is low and only slightly significant (data not shown). These results support the model of the physical measures of interference (in SC units) is conserved between sexes inwithin the same species.

@@

  For the normalized values in both M2 and M3, sex is a significant effect, increasing 
with males nrm.IFD in males.
  SKIVE*male is the only other consistently significant effect, which also increases the nrm.IFD measure (p = male-SKIVE p =  ).
For the mixed models of IFDs, sex is a significant effect for both raw and nrmIFD. (IFD.PER), sex is most significant effect (LRT: p = `r Q1.IFD.PER_M1_lmer.results$sex.results`)

@@

# Q1 Summary, consequences of sex differences in the recombination landscape

Our results confirm sex differences in recombination landscapes described in several other species and we note that these differences have distinct consequences on the potential patterns of genetic variation resulting from the distinct recombination landscapes.  
The terminal position of single foci and the greater distance between two foci on the same chromosome will have a consequence of larger sections of linked sites segregating together in male gametes compared to female gametes (Veller).

@@

The next section is meant to focus on the greater aspect of variation in mean MLH1 counts per cell the high recombining strains ( _musculus^PWD^_ , _musculus^SKIVE^_, and _molossinus^MSM^_) to the low recombing strains. The main objective of this section is to test for significant correlations between features of the recombination landscape and the evolution of mean MLH1 foci per cell.

In comparing the male specific single bivalent based metrics the first creteria for analysis is a significant differences between the high and low recombining groups and  the second step is testing for significant subspecies and strain effects in the same models which used mean MLH1 count per cell.

**Linear model M1 ** $$mouse \ average \ bivalent metric ~=~ subsp * strain + \varepsilon $$

**Linear model M2 ** $$mouse \ average \ bivalent metric ~=~ strain + \varepsilon $$

The literature provides basic predictions for the direction of the relationship the genome wide recombination rate and two aspects of the recombination landscape; SC length (positive) and crossover interference (negative). Following this logic we predict i) no strain effects for SC length or crossover interference within _domesticus_, ii) _musculus^PWD^_ will have greater SC length and weaker interference than _musculus^SKIVE^_, which will have longer SC and weaker interference compared to the other _musculus_ strains and iii) _molossinus^MSM^_ will have longer SC and weaker interference compared to _molossinus^MOLF^_.

@@

Given the differences in the DSB numbers between a subset of the high and low recombining strains we predict that there will also be differences in SC lengths proportional to the increase in the mean genome wide recombination rate. We compared mouse means of three SC based metrics; total SC, mean short bivalent length and mean long bivalent length. The long bivalent data set was isolated in the same manner as the short bivalent data set.

<!-- basic stats
`r length(OG.MLH1.merge.Skel$fileName)` total skelentonize measures were made from all of the MLH1 pachtyene cell images. X observations removed for having extra signal from other cells. `r length(unique( all.remove ) )` measures removed as potential outliers (by examining the category specific plots)<pattern for the single bivalent, pos correlation, The predicted pattern for SC length becomes more nuanced when the data are split up by Chromosome class. Mainly that 1CO are shorter in the high recombining strains then low recombining strains. This is likely due to the fact that more physically longer chromosomes have 1CO in the low recombining strains, which pushes the mean SC length up. Where as in the high recombining strains, physically longer chromosomes are more likely to be in the 2CO group.  This supports a general pattern of tighter clustering of SC lengths across chromosome classes in the high recombininggroup. They have a lower probability of 2COs below a certain SC length threshold.>  -->

@@

Confirming the basic predictions, there is a positive correlation between mean MLH1 foci per cell and total SC, however the variation in mean total SC does not completely predict the high and low recombining strains (Figure X). While for the total data set of mean total SC, the high recombining strains have significantly more SC area
(t.test; p = `r Q2tot.sc_t.test_High.Low$p.value`), only the _molossinus_ strains are significantly different between the high and low recombining strains (t.test mol p = `r Q2tot.sc_t.test_High.Low_Mol$p.value`). The _musculus_ are not significantly different across the groups (ttest; musc p= `r Q2tot.sc_t.test_High.Low_Musc$p.value`). Additionally, the means for the reduced bivalent datasets, short and long bivalents, are not significantly different between the high and low recombining strains (t.test; short p = `r Q2short.biv_t.test_High.Low$p.value` and long p = `r Q2long.biv_t.test_High.Low$p.value`).

@@

@@

For the linear model testing subspeices and strain (M1) and total SC as the dependant variable, the two subspecies effects are significant (glm; p= _Musculus_ `r sum.Q2.glm_mus.av_totSC_M1[,4][2]` and _Molossinus_ p= `r sum.Q2.glm_mus.av_totSC_M1[,4][3]`). Whereas the model with reduced bivalent means as the dependant variables, a number of subspeicies and strains effects reach significance, (glm; p < 0.05) they are not consistant across the sets of models tested, indecateing to some extent the chromation compaction evolution is decoupled from evolution in mean MLH1 foci per cell.

<!--
In the short bivalent data set, analyzed by M1, both subspecies are 
 (glm; p= _Musculus_ `r sum.glmQ2_short.biv_M1$coefficients[,4][2]` and _Molossinus_ p= `r sum.glmQ2_short.biv_M1$coefficients[,4][3]`). While the models for the reduced data set suggest that the subspecies with the high recombining strains have evolved longer SC lengths --
long bivalent set in M1 -- 
Musc `r sum.glmQ2_long.biv_M1$coefficients[,4][2]` 
MSM strain `r sum.glmQ2_long.biv_M1$coefficients[,4][7]`
<And the strain effect for G is significant (glm; G p =`r sum.glmQ2_short.biv_M1$coefficients[,4][4]`).>

@@

Unlike total SC length and single foci landscapes, our metrics of interference from 2COs are significant predictors of evolution in genome wide recombination rates. However, the pattern is in the opposite direction to our prediction; the high recombining strains have significantly stronger interference (instead of being predicted weaker interference). In the 2CO landscape, instead of observing weaker interference strength and two foci placed closer together on a single chromosomes, the two foci are spaced further apart.
